Bayesian Inference from Time Series: Information Loss and the Whittle Likelihood

### 2018-03-06 doc/ms.tex
New results and writing

\documentclass[a4paper, 12pt]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs} % For nicer tables
\usepackage{color}
\usepackage[left=2cm, right=2cm, bottom=3cm, top=2cm]{geometry}
\usepackage{graphicx}
\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{natbib}
\usepackage{transparent}

@@

Assume a quantity has been measured at $N$ equally-spaced time points
$\boldsymbol{t} = \{0, 1, 2, ..., N-1\}$, resulting in observations
$\boldsymbol{y} = \{y_0, y_1, ..., y_{N-1}\}$.
% Use GP framework

The whittle likelihood is
\begin{align}
p(\boldsymbol{y} | \boldsymbol{\theta})
    &\propto -\sum_{k=0}^{N/2 - 1}
         \left(S(\omega_k; \boldsymbol{\theta})
         + \frac{I(\omega_k; \boldsymbol{y})}
                {S(\omega_k; \boldsymbol{\theta})}\right)\label{eqn:whittle}
\end{align}
where $I(\omega_k; \boldsymbol{y})$ is the power spectrum of the data,
$S(\omega_k; \boldsymbol{\theta})$ is the power spectral density implied
by the model parameters, and the summation is over the angular frequencies
involved in the first half of the discrete fourier transform.
Equation~\ref{eqn:whittle} is equivalent to an exponential distribution
for the empirical power spectrum
$I(\omega_k; \boldsymbol{y})$ whose expectation is given by the
model power spectrum $S(\omega_k; \boldsymbol{\theta})$.

@@

\begin{table}[!ht]
\centering
\begin{tabular}{@{}l@{\hspace{3em}}l@{\hspace{2em}}l@{}}
\toprule
$N$         &       Likelihood      &  Mut. Inf. (nats) \\
\hline
20          &       Exact           & $1.930 \pm 0.055$ \\
50          &       Exact           & $2.391 \pm 0.056$ \\
100         &       Exact           & $2.605 \pm 0.053$ \\
200         &       Exact           &   \\
20          &       Whittle         & $1.491 \pm 0.055$ \\ 
50          &       Whittle         & $1.957 \pm 0.055$ \\
100         &       Whittle         & $2.052 \pm 0.055$ \\
200         &       Whittle         &   \\
\bottomrule
\end{tabular}
\caption{The mutual information between the data and the log-period, for
different numbers of datapoints $N$ and for different likelihoods (exact vs.
Whittle).
This is equal to the prior expected reduction in the entropy
in going from the prior to the posterior.\label{tab:oscillation_results}}
\end{table}

\begin{table}[!ht]
\centering
\begin{tabular}{@{}l@{\hspace{3em}}l@{}}
\toprule
$N$         &       $\Delta$ Mut. Inf. (nats) \\
\hline
20          &       $0.439 \pm 0.053$ \\
50          &       $0.434 \pm 0.056$ \\
100         &       $0.553 \pm 0.056$ \\
200         & \\
\bottomrule
\end{tabular}
\caption{The difference in mutual information between the exact and
Whittle likelihoods. This quantifies the size of the advantage of the exact
likelihood.
The error bars are smaller than what you might expect from
taking differences in Table~\ref{tab:oscillation_results}, due to
the common stream of datasets.\label{tab:oscillation_differences}}
\end{table}

@@

>>>>

\documentclass[a4paper, 12pt]{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs} % For nicer tables
\usepackage{color}
\usepackage[left=2cm, right=2cm, bottom=3cm, top=2cm]{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage{microtype}
\usepackage{natbib}
\usepackage{transparent}

@@

Assume a quantity has been measured at $N$ equally-spaced time points
$\boldsymbol{t} = \{0, 1, 2, ..., N-1\}$, resulting in observations
$\boldsymbol{y} = \{y_0, y_1, ..., y_{N-1}\}$.
Throughout this paper, I will specify all models in the framework of
Gaussian processes. Ultimately, this means the probability distribution for
the data $\boldsymbol{y}$ given some parameters $\boldsymbol{\theta}$ will
be a multivariate normal distribution whose mean vector $\boldsymbol{\mu}$
and covariance matrix $\mathbf{C}$ depend on $\boldsymbol{\theta}$:
\begin{align}
\boldsymbol{y} | \boldsymbol{\theta} &\sim
    \textnormal{Normal}\left(\boldsymbol{\mu}(\boldsymbol{\theta}),
                             \mathbf{C}(\boldsymbol{\theta})\right).
\end{align}
The full expression for this probability density is
\begin{align}
p(\boldsymbol{y} | \boldsymbol{\theta})
    &= \frac{1}{\sqrt{(2\pi)^N\det \mathbf{C}}}
        \exp\left[-\frac{1}{2}
                    (\boldsymbol{y} - \boldsymbol{\mu})^T
                    \mathbf{C}^{-1}
                    (\boldsymbol{y} - \boldsymbol{\mu})\right]
\end{align}
which also provides the exact likelihood function if we wish to infer
$\boldsymbol{\theta}$ from $\boldsymbol{y}$. The determinant and inverse
are typically expensive, though in some cases fast methods are available
\citep[e.g.][]{mackay, hodlr, celerite}.

The whittle likelihood is
\begin{align}
p(\boldsymbol{y} | \boldsymbol{\theta})
    &\propto -\sum_{k=0}^{N/2 - 1}
         \left[S(\omega_k; \boldsymbol{\theta})
         + \frac{I(\omega_k; \boldsymbol{y})}
                {S(\omega_k; \boldsymbol{\theta})}\right]\label{eqn:whittle}
\end{align}
where $I(\omega_k; \boldsymbol{y})$ is the power spectrum of the data,
$S(\omega_k; \boldsymbol{\theta})$ is the power spectral density implied
by the model parameters, and the summation is over the angular frequencies
involved in the first half of the discrete fourier transform.
Equation~\ref{eqn:whittle} is equivalent to an exponential distribution
for the empirical power spectrum
$I(\omega_k; \boldsymbol{y})$ whose expectation is given by the
model power spectrum $S(\omega_k; \boldsymbol{\theta})$.

@@

\begin{table}[!ht]
\centering
\begin{tabular}{@{}l@{\hspace{3em}}l@{\hspace{2em}}l@{}}
\toprule
$N$         &       Likelihood      &  Mut. Inf. (nats) \\
\hline
20          &       Exact           & $1.930 \pm 0.055$ \\
50          &       Exact           & $2.391 \pm 0.056$ \\
100         &       Exact           & $2.605 \pm 0.053$ \\
200         &       Exact           & $2.830 \pm 0.054$ \\
20          &       Whittle         & $1.491 \pm 0.055$ \\ 
50          &       Whittle         & $1.957 \pm 0.055$ \\
100         &       Whittle         & $2.052 \pm 0.055$ \\
200         &       Whittle         & $2.158 \pm 0.061$  \\
\bottomrule
\end{tabular}
\caption{The mutual information between the data and the log-period, for
different numbers of datapoints $N$ and for different likelihoods (exact vs.
Whittle).
This is equal to the prior expected reduction in the entropy
in going from the prior to the posterior.\label{tab:oscillation_results}}
\end{table}

\begin{table}[!ht]
\centering
\begin{tabular}{@{}l@{\hspace{3em}}l@{}}
\toprule
$N$         &       $\Delta$ Mut. Inf. (nats) \\
\hline
20          &       $0.439 \pm 0.053$ \\
50          &       $0.434 \pm 0.056$ \\
100         &       $0.553 \pm 0.056$ \\
200         &       $0.671 \pm 0.062$ \\
\bottomrule
\end{tabular}
\caption{The difference in mutual information between the exact and
Whittle likelihoods. This quantifies the size of the advantage of the exact
likelihood.
The error bars are smaller than what you might expect from
taking differences in Table~\ref{tab:oscillation_results}, due to
the common stream of datasets.\label{tab:oscillation_differences}}
\end{table}

@@

\bibitem[Gibbs and MacKay(1997)]{mackay}
MacKay, D., and M. N. Gibbs. ``Efficient implementation of Gaussian processes.'' Neural Computation (1997).
Available at \url{http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.48.8224&rep=rep1&type=pdf}
(as of 2018-03-07).
